%%% Document Formatting
\documentclass[12pt,fleqn]{article}
\usepackage[a4paper,
            bindingoffset=0.2in,
            left=0.75in,
            right=0.75in,
            top=0.8in,
            bottom=0.8in,
            footskip=.25in]{geometry}
\setlength\parindent{10pt} % No indent

%%% Imports
% Mathematics
\usepackage{amsmath} % Math formatting
\numberwithin{equation}{section} % Number equation per section
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\sgn}{sgn}



\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{hypothesis}{Hypothesis}
\newtheorem{proof}{Proof}
\newtheorem{lemma}{Lemma}
\newtheorem{example}{Example}

\usepackage{amsmath}
\usepackage{amsfonts} % Math fonts
\usepackage{amssymb} % Math symbols
\usepackage{mathtools} % Math etc.
\usepackage{slashed} % Dirac slash notation
\usepackage{cancel} % Cancels to zero
\usepackage{empheq}
\usepackage{breqn}

\newcommand{\expect}{\mathbb{E}}


% Visualization
\usepackage{graphicx} % for including images
\graphicspath{ {} } % Path to graphics folder
\usepackage{tikz}



%%% Formating
\usepackage{hyperref} % Hyperlinks
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    pdftitle={Overleaf Example},
    pdfpagemode=FullScreen,
    }
\urlstyle{same}

\usepackage{mdframed} % Framed Enviroments
\newmdenv[ 
  topline=false,
  bottomline=false,
  skipabove=\topsep,
  skipbelow=\topsep
]{sidework} %% Side-work

\usepackage{lipsum} % Lorem Ipsum example text

%%%%% ------------------ %%%%%
%%% Title
\title{Lectures Notes from Beg Rohu 2025}
\author{Clark Miyamoto (cm6627@nyu.edu)}
\date{\today}
\begin{document}

\maketitle
\begin{abstract}
	A set of lecture note from the Beg Rohu Summer School in 2025. If any of the other participants would like to collaborate on these notes, please reach out to me!
\end{abstract}

\tableofcontents
\newpage

\part{Agliari: Information Processing in Hebbian Networks}
\input{agliari}
\newpage

\part{Olivier Dauchot: Swarm Robotics as Smart Active Matter}
\input{dauchot}
\newpage

\part{Julia Kempe: Synthetic Data}
\input{kempe}
\newpage

\part{Jean R\'emi King: Geometry of Thought}
\input{king}
\newpage

\part{Yann LeCun: Self Supervised Models}
\input{lecun}
\newpage

\part{St\'ephane Mallat: Score Based Diffusion \& Renormalization Group Flow}
\input{mallat}
\newpage

\part{Marc M\'ezard: Statistical Physics of Generative Diffusion}
\input{mezard}
\newpage

\part{Arvind Murugan: Learning without Neurons}
\input{murugan}
\newpage

\part{Etc.}

\section{People Presentations} 
\begin{enumerate}
	\item Paolo Balioni, working on feature learning in Bayesian Neural Networks. Also works on MCMC methods.
	\item Claudian Merger. Mappping normalizing flows to $\phi^n$ theory.
\end{enumerate}

\section{Acknowledgements}
Thank you to Prof. David Hogg \& the Flatiron Institute for providing funding for this trip!





\end{document}